Some examples in extended logic counting classes

### 2016-12-28 LICS/preliminaries.tex
19 dic
Avance en sección 3.

\marcelo{Agregar las definiciones de LFP y PFP}
\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a relational signature $\R$,  function $f$ is said to be {\em defined over $\R$} (or just {\em over $\R$}) if $f$ is a total function from $\{\enc(\A) \mid \A \in \ostr[\R]\}$ to $\nat$. Moreover, a set $\CC$ of functions is said to be a {\em function complexity class} if every $f \in \CC$ is a function defined over some relational signature $\R$.

\marcelo{Ya no consideramos clases de estructuras, hay que revisar que esta seccion tenga una notacion consistente}
\martin{done}

\marcelo{Agregar las definiciones de las clases usuales de funciones: FP, \#P, ...}
\martin{done}

We define the following function complexity classes: $\fp$ is defined as the class of functions computable in polynomial time. $\shp$ is defined as the class of functions that on an input are equal to the number of accepting paths of an $\np$ machine on that input. $\spp$ is defined as the class of functions that on an input are equal to the number of different outputs of an $\np$ transducer on that input. $\fpspace$ is the class of functions computable in polynomial space. $\nfpspace$ is the class of single-exponentially bounded functions computable in polynomial space.

>>>>

\cristian{Agregar que la igualdad $=$ esta en nuestra lógica.}

\cristian{Agregar acá que usamos las abreviaciones de $<$ o $\subset$.}

\marcelo{Agregar las definiciones de LFP y PFP}

\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a relational signature $\R$,  function $f$ is said to be {\em defined over $\R$} (or just {\em over $\R$}) if $f$ is a total function from $\{\enc(\A) \mid \A \in \ostr[\R]\}$ to $\nat$. Moreover, a set $\CC$ of functions is said to be a {\em function complexity class} if every $f \in \CC$ is a function defined over some relational signature $\R$.

We define the following function complexity classes: $\fp$ is defined as the class of functions computable in polynomial time. $\shp$ is defined as the class of functions that on an input are equal to the number of accepting paths of an $\np$ machine on that input. $\spp$ is defined as the class of functions that on an input are equal to the number of different outputs of an $\np$ transducer on that input. $\fpspace$ is the class of functions computable in polynomial space. $\shpspace$ is defined analogously to $\shp$ but in polynomial space. $\nfpspace$ is the class of single-exponentially bounded functions computable in polynomial space.

### 2016-12-28 LICS/saluja.tex

%\begin{enumerate}
%	\item Introduce the classes $\eqso(\LL)$.
%	
%	\item Study the properties of $\eqso(\LL)$.
%	\begin{itemize}
%		\item If $\LL$ is closed under conjunction, then binary product can be defined in $\eqso(\LL)$.
%		
%		\item If $\LL$ is closed under XXX, then the first order product can be defined in $\eqso(\LL)$.
%		
%		\item Every formula in $\eqso(\LL)$ can be rewrite in \emph{sum normal form}.
%	\end{itemize}
%	
%	\item Introduce Saluja classes.
%	
%	\item Show the connection with Saluja classes.
%	\begin{itemize}
%		\item New classes are properly contained.
%		\item Old classes are properly contained/equivalent in new classes. 
%	\end{itemize}
%	
%	\item Study the complexity of new classes.
%	\begin{itemize}
%		\item $\eqso(\Sigma_0)$ is in FP.
%		\item $\eqso(\Sigma_1)$ is $\#P$ easy.
%		\item $\eqso(\Sigma_1)$ has free pass.
%	\end{itemize}
%	
%	\item Closure properties.
%	\begin{itemize}
%		\item $\eqso(\Sigma_1)$ is not closed under substraction.
%		\item $\eqso(\Sigma_1)$ is not closed under minus 1.
%	\end{itemize}
%	
%\end{enumerate}

@@

In the next result, we show that $\eqso$ also defined a finite hierarchy similar than in~\cite{SalujaST95} that we called the $\eqso$-hierarchy.
\begin{proposition}
\[
\QE{0} \; \subsetneq \; \QE{1} \; \subsetneq \; \QU{1} \; \subsetneq \; \QE{2} \; \subsetneq \; \QU{2} \; = \; \eqso{\fo} 
\]
\end{proposition}
A natural question until this point is what is the connection between $\E{i}$ and $\U{i}$ and their corresponding classes $\QE{i}$ and $\QU{i}$. Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ or $\U{i}$ is equivalent to $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ which is in $\QE{i}$ or $\QU{i}$, respectively. 
Then it left to know whether the containment between the sharp and quantitative classes is strict or not. 
For this, we study first whether each formula $\alpha$ in $\eqso(\LL)$ can be transformed in a \emph{prefix normal form} where all sums quantifiers are restricted to the beginning of the formula. 
Formally, we say that a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\Sigma$-prenex normal form} (or just prenex normal form) if $\alpha$ is of the form:
\[
\alpha := \sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})
\]
where $\bar{X}$ and $\bar{x}$ is a sequence of $0$ or more second and first order variables, respectively.
In particular, note that a boolean formula is always in $\Sigma$-prenex normal form.
Then we say that a formula $\alpha$ is in \emph{sum normal form} if $\alpha$ is of the form $c + \Sigma_{i=1}^n \alpha_i$ where $c$ is a non-negative constant and each $\alpha_i$ is in $\Sigma$-prenex normal form.
Next we show that each formula in $\eqso(\LL)$ can be converted in \emph{sum normal form} but not always in prenex normal form.
\begin{theorem}
	Every formula in $\eqso(\LL)$ can be rewritten in \emph{sum normal form}. Furthermore, there exists a formula $\alpha \in \loge{1}$ such that there does not exist a formula in $\Sigma$-prenex normal form in $\QE{1}$ equivalent to $\alpha$.
\end{theorem}

@@

The previous results gives the connection the connection between the hierarchy in \cite{SalujaST95} and the $\eqso$-hierarchy. 
Indeed, if $\LL$ contains $\logu{1}$ we will have that $\sh{\LL}$ is equivalent to $\eqso(\LL)$ since each formula in $\eqso(\LL)$ can be converted in $\Sigma$-prefix normal form and, therefore, it is equivalent to a formula in $\sh{\LL}$. 
Unfortunately, this is not the case for $\loge{0}$ and $\loge{1}$ as the following result shows.
\begin{proposition}
	The classes $\E{0}$ and $\E{1}$ are strictly contained in $\QE{0}$ and $\QE{1}$, respectively. Moreover, the classes $\U{1}$, $\E{2}$, and $\U{2}$ are equivalent with $\QU{1}$, $\QE{2}$, and $\QU{2}$, respectively.
\end{proposition}
The previous result shows that the classes $\QE{i}$ and $\QU{i}$ are more robust than the classes $\E{i}$ and $\U{i}$: they are closed under binary and sum quantifiers but the other not necessarily.

@@

By the previous result, we know that functions in the $\eqso$ hierarchy are not closed under subtraction. Then, a natural restriction to this question is to ask whether these classes are closed under subtraction by one, namely, if $\CC$ is a class of functions and $f \in \CC$, is $f-1 \in \CC$ where $1$ is the constant function that outputs $1$ for every structure. 
We do not know $\E{1}$ is closed under subtraction by one. However, if we extend $\logex{1}$ with $\fo$ predicates we can show that this new fragment is closed under subtraction by one.
\begin{theorem} \label{sigmafo-minusone}
	$\eqso(\logex{1})$ is closed under substraction by one.
\end{theorem}

>>>>

%!TEX root = syntactic.tex

@@

In the next result, we show that $\eqso$ also defined a finite hierarchy similar than in~\cite{SalujaST95} that we called the $\eqso$-hierarchy.
\begin{proposition}
\[
\QE{0} \; \subsetneq \; \QE{1} \; \subsetneq \; \QU{1} \; \subsetneq \; \QE{2} \; \subsetneq \; \QU{2} \; = \; \eqso(\fo)
\]
\end{proposition}
A natural question until this point is what is the connection between $\E{i}$ and $\U{i}$ and their corresponding classes $\QE{i}$ and $\QU{i}$. Clearly, we have that $\E{i} \subseteq \QE{i}$ and $\U{i} \subseteq \QU{i}$. Indeed, each formula $\varphi(\bar{X}, \bar{x})$ in $\E{i}$ or $\U{i}$ is equivalent to $\sa{\bar X} \sa{\bar x} \varphi(\bar{X}, \bar{x})$ which is in $\QE{i}$ or $\QU{i}$, respectively. 
Then it is left to know whether the containment between the sharp and quantitative classes is strict or not. 
For this, we study first whether each formula $\alpha$ in $\eqso(\LL)$ can be transformed in a \emph{prefix normal form} where all sums quantifiers are restricted to the beginning of the formula. 
Formally, we say that a formula $\alpha$ in $\eqso(\LL)$ is in \emph{$\Sigma$-prenex normal form} (or just prenex normal form) if $\alpha$ is of the form:
\[
\alpha := \sa{\bar{X}} \sa{\bar{x}} \varphi(\bar{X}, \bar{x})
\]
where $\bar{X}$ and $\bar{x}$ is a sequence of $0$ or more second and first order variables, respectively.
In particular, note that a boolean formula is always in $\Sigma$-prenex normal form.
Then we say that a formula $\alpha$ is in \emph{sum normal form} if $\alpha$ is of the form $c + \Sigma_{i=1}^n \alpha_i$ where $c$ is a non-negative constant and each $\alpha_i$ is in $\Sigma$-prenex normal form.
Next we show that each formula in $\eqso(\LL)$ can be converted in \emph{sum normal form} but not always in prenex normal form.
\begin{theorem}
	Every formula in $\eqso(\LL)$ can be rewritten in \emph{sum normal form}. Furthermore, there exists a formula $\alpha \in \loge{1}$ such that there does not exist a formula in $\Sigma$-prenex normal form in $\QE{1}$ equivalent to $\alpha$.
\end{theorem}

@@

The previous results gives the connection between the hierarchy in \cite{SalujaST95} and the $\eqso$-hierarchy. 
Indeed, if $\LL$ contains $\logu{1}$ we will have that $\sh{\LL}$ is equivalent to $\eqso(\LL)$ since each formula in $\eqso(\LL)$ can be converted in $\Sigma$-prefix normal form and, therefore, it is equivalent to a formula in $\sh{\LL}$. 
Unfortunately, this is not the case for $\loge{0}$ and $\loge{1}$ as the following result shows.
\begin{proposition}
	The classes $\E{0}$ and $\E{1}$ are strictly contained in $\QE{0}$ and $\QE{1}$, respectively. Moreover, the classes $\U{1}$, $\E{2}$, and $\U{2}$ are equivalent with $\QU{1}$, $\QE{2}$, and $\QU{2}$, respectively.
\end{proposition}
The previous result shows that the classes $\QE{i}$ and $\QU{i}$ are more robust than the classes $\E{i}$ and $\U{i}$: they are closed under binary and sum quantifiers but the other not necessarily.

@@

By the previous result, we know that functions in the $\eqso$ hierarchy are unlikely to be closed under subtraction. Then, a natural restriction to this question is to ask whether these classes are closed under subtraction by one, namely, if $\CC$ is a class of functions and $f \in \CC$, is $f-1 \in \CC$ where $1$ is the constant function that outputs $1$ for every structure. 
We do not know $\E{1}$ is closed under subtraction by one. However, if we extend $\logex{1}$ with $\fo$ predicates we can show that this new fragment is closed under subtraction by one.
\begin{theorem} \label{sigmafo-minusone}
	$\eqso(\logex{1})$ is closed under substraction by one.
\end{theorem}
